The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-10-01 paper.qmd
fixed more -ize

When we use "uncertainty visualization" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
Graphics, just like statistics, are not defined by their input data. 
A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. 
Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special. 
Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty. 
Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.

@@

This interpretation of uncertainty indicates that the uncertainty is not a variable of importance in of itself. 
Uncertainty is metadata about our statistic that is required for valid inference.
This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. 
This aspect of uncertainty visualization makes it a uniquely difficult problem.

@@

Some authors recognise inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. 
@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualization. 
This makes it a little difficult to ensure uncertainty is always included.
For example, if we have a visualisation that shows an average, we would need to identify if the signal suppression should be performed using the sampling variance or the sample variance [@Hofman2020].
The distribution we use depends on the inferential statistic but until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated.
This means the ideal uncertainty visualisation should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.

@@

Risk-aversion is another secondary effect of uncertainty visualization that is used to evaluate uncertainty visualisations [@Hullman2019]. 
Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
Rational agents *by definition* should *ignore* uncertainty information. 
Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
The only case where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.

@@

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
@Padilla2017 seem to be aware of the issues in the visualisation and are using the user study to show that current government communications are not the best choice when communicating hurricane risk, as it is unreasonable to assume that "more likely to be hit" should automatically translate to "more damage". 
While this is useful to highlight, the authors end up caught by their own trap, as answering the questions in their second study correctly required participants to assume an area being more likely to be hit *does* translate to receiving more damage.
The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

@@

### Comparing to hypothesis tests
The most obvious way to evaluate uncertainty visualizations is to compare the visualisations to statistical tests. 
If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test. Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation. 
The lineup protocol is a good example of this approach.
Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. 
In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. 
This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy. 
The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is "the expected" and we are looking for "the unexpected".
This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.

>>>>

When we use "uncertainty visualation" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
Graphics, just like statistics, are not defined by their input data. 
A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. 
Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special. 
Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty. 
Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.

@@

This interpretation of uncertainty indicates that the uncertainty is not a variable of importance in of itself. 
Uncertainty is metadata about our statistic that is required for valid inference.
This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. 
This aspect of uncertainty visualation makes it a uniquely difficult problem.

@@

Some authors recognise inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. 
@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualisation. 
This makes it a little difficult to ensure uncertainty is always included.
For example, if we have a visualisation that shows an average, we would need to identify if the signal suppression should be performed using the sampling variance or the sample variance [@Hofman2020].
The distribution we use depends on the inferential statistic but until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated.
This means the ideal uncertainty visualisation should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.

@@

Risk-aversion is another secondary effect of uncertainty visualisation that is used to evaluate uncertainty visualisations [@Hullman2019]. 
Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualisation experiments [@wu2023rational].
However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
Rational agents *by definition* should *ignore* uncertainty information. 
Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
The only case where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.

@@

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
@Padilla2017 seem to be aware of the issues in the visualisation and are using the user study to show that current government communications are not the best choice when communicating hurricane risk, as it is unreasonable to assume that "more likely to be hit" should automatically translate to "more damage". 
While this is useful to highlight, the authors end up caught by their own trap, as answering the questions in their second study correctly required participants to assume an area being more likely to be hit *does* translate to receiving more damage.
The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualation based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

@@

### Comparing to hypothesis tests
The most obvious way to evaluate uncertainty visualations is to compare the visualisations to statistical tests. 
If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test. Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation. 
The lineup protocol is a good example of this approach.
Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. 
In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. 
This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy. 
The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is "the expected" and we are looking for "the unexpected".
This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.
